---
title: "Example - Export to Darwin Core format"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(dplyr)
```

This demonstrates exporting the example Level 3 output dataset to the Darwin Core Archive format for packaging and publication to biodiversity repositories (eg. GBIF, OBIS, Living Atlases). The format is described in the Darwin Core [text guide](https://dwc.tdwg.org/text/). This example uses the OBIS ENV-DATA approach, described [here](https://obis.org/manual/dataformat/).

## Data In

The data used here are associated with Braun CD, Gaube P, Sinclair-Taylor TH, Skomal GB, Thorrold SR (2019) Mesoscale eddies release pelagic sharks from thermal constraints to foraging in the ocean twilight zone. Proc Natl Acad Sci U S A 116:17187–17192 (doi: [10.1073/pnas.1903067116](https://doi.org/10.1073/pnas.1903067116)), and the data are archived at DataOne (doi: [10.24431/rw1k329](https://doi.org/10.24431/rw1k329)).

Reading the data from the files which are formatted to the metadata templates:

```{r eval = TRUE, echo = TRUE}
deployment_metadata <- read.csv("../braun_blues_deployment-metadata.csv")
device_metadata <- read.csv("../braun_blues_device-metadata.csv")
obs_data <- read.csv("../data_level3/blue_sharks_level3.csv")
```

## Darwin Core entities

The data is spread across a number of files which are zipped into the archive:

* __event.csv__ the Event Core outlines major project milesones such as capture, deployment, and the deployment completion
* __occurrence.csv__ the Occurrence Extension contains each occurrence record, usually a HumanObservation which is associated with the initial capture and biota measurements, then a series of MachineObservations associated with the tag and sensor information.
* __extendedMeasurementOrFact.csv__ the Extended Measurement or Fact Extension contains any measurement or information related to any Event or Occurrence, eg. sensor metadata
* __meta.xml__ provides a machine readable document describing the structure of the archive
* __eml.xml__ contains standard scientific publication metadata expressed in Ecological Modelling Language

### Event Core

3 key events in this dataset can be mapped from the __deployment-metadata__:

* Deployment Start / capture
* Tag Deployment
* Deployment End

```{r eval = TRUE, include = FALSE}
starteventdf <- deployment_metadata %>% mutate(eventID = paste0(deploymentID,":start")
                                          , organismID = organismID
                                          , eventRemarks = "Deployment Start"
                                          , eventDate = deploymentDateTime
                                          , decimalLatitude = deploymentLatitude
                                          , decimalLongitude = deploymentLongitude) %>% 
                                   select(eventID, organismID, eventRemarks, eventDate, decimalLatitude, decimalLongitude)

deployeventdf <- deployment_metadata %>% mutate(eventID = deploymentID
                                          , organismID = organismID
                                          , eventRemarks = "Deployment"
                                          , eventDate = paste0(deploymentDateTime,"/",detachmentDateTime) 
                                          , decimalLatitude = as.numeric(NA)
                                          , decimalLongitude= as.numeric(NA)) %>% 
                                   select(eventID, organismID, eventRemarks, eventDate, decimalLatitude, decimalLongitude)

endeventdf <- deployment_metadata %>% mutate(eventID = paste0(deploymentID,":end")
                                          , organismID = organismID
                                          , eventRemarks = "Deployment End"
                                          , eventDate = detachmentDateTime
                                          , decimalLatitude = detachmentLatitude
                                          , decimalLongitude = detachmentLongitude) %>% 
                                   select(eventID, organismID, eventRemarks, eventDate, decimalLatitude, decimalLongitude)
eventdf <- bind_rows(starteventdf,deployeventdf,endeventdf)
```

These event records are linked to from occurrence and measurement records.

```{r eval = TRUE}
eventdf %>% filter(grepl("160424_2016_106744",eventID))
```

### Occurrence Extension

Biodiversity infrastructures index by occurrence records, denoting species-location-time, and manage them by a semi-controlled vocabulary in the [basisOfRecord term](https://dwc.tdwg.org/terms/#dwc:basisOfRecord) which delineates HumanObservations, MachineObservations, PreservedSpecimen etc.

The archive will record a HumanObservation record which aligns with the animal capture/deployment start event, and will link to observed biotic measurements. 
MachineObservation occurrence records are features recorded by the tag with spatial and temporal components.

```{r eval = TRUE, include = FALSE}
#eventID, occurrenceID, eventDate, basisOfRecord, scientificName, vernacularName, scientificNameID, decimalLatitude, decimalLongitude, sex, 

humanobsdf <- deployment_metadata %>%  mutate(eventID = paste0(deploymentID,":start")
                                               , occurrenceID = paste0(deploymentID,":start")
                                               , eventDate = deploymentDateTime
                                               , basisOfRecord = "HumanObservation"
                                               , organismID = organismID
                                               , decimalLatitude = deploymentLatitude
                                               , decimalLongitude = deploymentLongitude
                                               , scientificName = scientificName
                                               , vernacularName = commonName
                                               , scientificNameID = scientificNameSource
                                               , sex = organismSex
                                               , reproductiveCondition = organismAgeReproductiveClass) %>% 
                                          select(eventID 
                                               , occurrenceID
                                               , eventDate
                                               , basisOfRecord
                                               , organismID
                                               , decimalLatitude
                                               , decimalLongitude
                                               , scientificName
                                               , vernacularName
                                               , scientificNameID
                                               , sex
                                               , reproductiveCondition)

machineobsdf <- obs_data %>% mutate(eventID = id
                                  , occurrenceID = paste0(id,":",row_number())
                                  , eventDate = date
                                  , basisOfRecord = "MachineObservation"
                                  , decimalLatitude = lat
                                  , decimalLongitude = lon
                                  , scientificName = 'Prionace glauca') %>% 
                           select(eventID 
                                  , occurrenceID
                                  , eventDate
                                  , basisOfRecord
                                  , decimalLatitude
                                  , decimalLongitude
                                  , scientificName)
occurrencesdf <- bind_rows(humanobsdf,machineobsdf) 
```

```{r eval = TRUE}
head(occurrencesdf %>% filter(grepl("160424_2016_106744",eventID)),3)
```

### Measurement or Fact Extension

This extension contains measurements and data that are not defined in the DarwinCore terms, have different cardinality to the occurrence or event records (ie many acceleromter records) and usually without a spatial component. 

In this example this includes all of the tag/device metadata, some of the animal measurements, and the detachment information. Each term should have a definition URI. 

```{r eval = TRUE, include = FALSE}

# mof ext fields: eventID, occurrenceID, measurementType, measurementTypeID, measurementValue, measurementValueID, measurementUnit, measurementUnitID

trapping_df <- deployment_metadata %>% mutate(eventID = paste0(deploymentID,":start")
                                                , occurrenceID = paste0(deploymentID,":start")
                                                , measurementType = "trappingMethodDetails"
                                                , measurementTypeID = ""
                                                , measurementValue = as.character(trappingMethodDetails)) %>%
                                           select(eventID, occurrenceID, measurementType, measurementTypeID, measurementValue)

attachment_df <- deployment_metadata %>% mutate(eventID = paste0(deploymentID,":start")
                                                , occurrenceID = paste0(deploymentID,":start")
                                                , measurementType = "attachmentMethod"
                                                , measurementTypeID = "https://github.com/tagbase/tagbase/blob/master/eTagMetadataInventory.csv#L26"
                                                , measurementValue = attachmentMethod) %>%
                                           select(eventID, occurrenceID, measurementType, measurementTypeID, measurementValue)

instrument_id_df <- deployment_metadata %>% mutate(eventID = deploymentID
                                                , occurrenceID = ""
                                                , measurementType = "instrumentID"
                                                , measurementTypeID = "http://vocab.nerc.ac.uk/collection/W07/current/IDEN0008/"
                                                , measurementValue = instrumentID) %>%
                                           select(eventID, occurrenceID, measurementType, measurementTypeID, measurementValue)

ptt_df  <- deployment_metadata %>% mutate(eventID = deploymentID
                                                , occurrenceID = ""
                                                , measurementType = "PTT"
                                                , measurementTypeID = "https://github.com/tagbase/tagbase/blob/master/eTagMetadataInventory.csv#L12"
                                                , measurementValue = as.character(ptt)) %>%
                                           select(eventID, occurrenceID, measurementType, measurementTypeID, measurementValue)


# for each instrument, add instrumentModel, instrumentManufacturer, trackingDevice, uplinkInterval
instrument_model_df <- deployment_metadata %>% mutate(eventID = deploymentID
                                                , occurrenceID = ""
                                                , measurementType = "instrumentModel"
                                                , measurementTypeID = "https://github.com/tagbase/tagbase/blob/master/eTagMetadataInventory.csv#L6"
                                                , measurementValue = "SPOT258G") %>%
                                           select(eventID, occurrenceID, measurementType, measurementTypeID, measurementValue)

instrument_manu_df <- deployment_metadata %>% mutate(eventID = deploymentID
                                                , occurrenceID = ""
                                                , measurementType = "instrumentManufacturer"
                                                , measurementTypeID = "http://vocab.nerc.ac.uk/collection/W07/current/IDEN0012/"
                                                , measurementValue = "Wildlife Computers") %>%
                                           select(eventID, occurrenceID, measurementType, measurementTypeID, measurementValue)

detachment_df <- deployment_metadata %>% mutate(eventID = paste0(deploymentID,":end")
                                                , occurrenceID = paste0(deploymentID,":end")
                                                , measurementType = "detachmentDetails"
                                                , measurementTypeID = ""
                                                , measurementValue = detachmentDetails) %>%
                                           select(eventID, occurrenceID, measurementType, measurementTypeID, measurementValue)

mofdf <- bind_rows(trapping_df,attachment_df,instrument_id_df,ptt_df,instrument_model_df,instrument_manu_df,detachment_df)
```
```{r eval = TRUE}
head(mofdf %>% filter(grepl("160424_2016_106744",eventID)))
```

## Final packaging 

Each file is packaged into a zip with metadata files meta.xml and eml.xml.

```{r eval = FALSE, echo = TRUE}
write.csv(eventdf,"event.csv",row.names = FALSE)
write.csv(occurrencesdf,"occurrence.csv",row.names = FALSE)
write.csv(mofdf,"mof.csv",row.names = FALSE)
zip("braun-blueshark-dwca.zip",files=c("event.csv","occurrence.csv","mof.csv","eml.xml","meta.xml"))
```
