---
title: "Blue shark standardization example"
output: github_document
---
  
```{r setup, eval = TRUE, message = FALSE}
library(sp)
library(tidyverse)
library(raster)
library(lubridate)
library(fields)
library(foieGras)

```


This example demonstrates the standardization process using an example set of blue shark satellite tag data. The data used here are associated with Braun CD, Gaube P, Sinclair-Taylor TH, Skomal GB, Thorrold SR (2019) Mesoscale eddies release pelagic sharks from thermal constraints to foraging in the ocean twilight zone. Proc Natl Acad Sci U S A 116:17187–17192 (doi: [10.1073/pnas.1903067116](https://doi.org/10.1073/pnas.1903067116)), and the data are archived at DataOne (doi: [10.24431/rw1k329](https://doi.org/10.24431/rw1k329)).

## Raw to Level 1
Preparing data for Level 1. Conversion from format output by Wildlife Computers Portal to standardized set of Level 1 position data.

```{r level1, eval = TRUE}
## list of raw Wildlife Computers tag data files
fList <- list.files('./wildlife_computers_raw/', full.names = TRUE, recursive = TRUE)
loc_files <- fList[grep('Locations', fList)]

## this metadata is modified from the original metadata at the DataOne repo to match the proposed standardization templates
meta <- read.table('braun_blues_deployment-metadata.csv', sep = ',', header = TRUE, stringsAsFactors = FALSE)
meta$trackStartTime <- lubridate::parse_date_time(meta$trackStartTime, orders = 'YmdHMS', tz = 'UTC')
meta$trackEndTime <- lubridate::parse_date_time(meta$trackEndTime, orders = 'YmdHMS', tz = 'UTC')

for (i in 1:length(loc_files)){
  
  ## read and format raw location data
  track <- read.table(loc_files[i], sep = ',', header = TRUE)
  track$Date <- lubridate::parse_date_time(track$Date, orders = c('HMS dbY', 'mdy HM', 'Ymd HMS'), tz = 'UTC')
  
  ## identify metadata row
  meta_idx <- which(meta$ptt %in% track$Ptt[1])
  
  ## identify start stop timestamps
  start <- meta$trackStartTime[meta_idx]
  stop <- meta$trackEndTime[meta_idx]
  
  ## filter raw data to data between track start/end times
  track <- track[which(track$Date >= start & track$Date <= stop),]

  ## add missing columns to dataframe
  track$instrumentID <- meta$instrumentID[meta_idx]
  track$deploymentID <- meta$deploymentID[meta_idx]
  track$organismID <- meta$organismID[meta_idx]
  
  ## existing column names to standardized column names
  names(track)[which(names(track) == 'Date')] <- 'time'
  names(track)[which(names(track) == 'Latitude')] <- 'latitude'
  names(track)[which(names(track) == 'Longitude')] <- 'longitude'
  names(track)[which(names(track) == 'Quality')] <- 'argosLC'
  names(track)[which(names(track) == 'Error.radius')] <- 'argosErrorRadius'
  names(track)[which(names(track) == 'Error.Semi.major.axis')] <- 'argosSemiMajor'
  names(track)[which(names(track) == 'Error.Semi.minor.axis')] <- 'argosSemiMinor'
  names(track)[which(names(track) == 'Error.Ellipse.orientation')] <- 'argosOrientation'

  ## time to ISO8601 format
  track$time <- lubridate::format_ISO8601(track$time)
  
  ## formatting to complete Level 1 coercion
  track <- track[,c('instrumentID','deploymentID','organismID', 'time', 'latitude', 'longitude', 'argosLC', 'argosErrorRadius', 'argosSemiMajor', 'argosSemiMinor', 'argosOrientation')]

  ## write level 1 data results
  write.table(track, paste('./data_level1/', meta$deploymentID[meta_idx], '_L1_data.csv', sep=''), sep=',', col.names = TRUE, row.names = FALSE)
  
  print(paste(meta$deploymentID[meta_idx], 'complete.', sep = ' '))
  
}

head(track)

```


## Level 1 -> 2

This step converts from "raw" location data to a more filtered and QC'd version of the "raw" data. Note that only very minor filtering is done at this stage to remove erroneous positions, but the location data is still the "raw" data.

```{r 1to2, eval = TRUE}
## list of all level 1 data files
loc_files <- list.files('./data_level1/', full.names = TRUE, recursive = TRUE)

for (i in 1:length(loc_files)){
  
  ## read and format raw, level 1 location data
  track <- read.table(loc_files[i], sep = ',', header = TRUE, stringsAsFactors = F)
  track$time <- lubridate::parse_date_time(track$time, orders = 'YmdHMS', tz = 'UTC')
  
  ## identify metadata row
  meta_idx <- which(meta$deploymentID %in% track$deploymentID[1])
  
  ## remove Argos quality Z locations
  track <- track[which(track$argosLC != 'Z' & track$argosLC != ' '),]
  
  ## removes duplicate POSIXct timestamps (not duplicate DAYS)
  ## could get smarter here, for example, by taking the better LC where times are duplicated
  track <- track[which(!duplicated(track$time)),]

  ## create spatial points for coercion to object of class "trip"
  tr <- track
  coordinates(tr) <- ~longitude + latitude
  proj4string(tr) <- CRS("+proj=longlat +ellps=WGS84 +datum=WGS84")

  ## then coerce to trip object for filtering
  tr <- trip::trip(tr, c('time','deploymentID'))
  
  ## filter at 4 m/s or 14.4 km/hr
  ## speed filter decision based on Fig S1 in Braun CD, Galuardi B, Thorrold SR (2018) HMMoce: An R package for improved geolocation of archival-tagged fishes using a hidden Markov method. Methods Ecol Evol 9:1212–1220
  sf2 <- trip::speedfilter(tr, max.speed = 14.4)
  
  ## subset track based on logical output of speed filter
  track <- track[sf2,]
  
  ## manually remove a spurious location that filters do not catch (note added to deployment-metadata QC section). this could easily be done in bulk using other methods
  if (track$deploymentID[1] == '160424_2015_141261') track <- track[which(track$time != as.POSIXct('2015-10-14 17:18:20', tz='UTC')),] 
  
  ## coerce timestamps back to ISO
  track$time <- lubridate::format_ISO8601(track$time)

  ## write level 2 data results
  write.table(track, paste('./data_level2/', meta$deploymentID[meta_idx], '_L2_data.csv', sep=''), sep=',', col.names = TRUE, row.names = FALSE)
  
  print(paste(meta$deploymentID[meta_idx], 'complete.', sep = ' '))
 
}

head(track)

```


## Level 2 -> 3

This step converts cleaned, raw data to a standardized set of location data interpolated to a regular temporal scale.

```{r level2to3, eval = TRUE, cache = TRUE}

loc_files <- list.files('./data_level2/', full.names = TRUE, recursive = TRUE)

for (i in 1:length(loc_files)){
  
  track <- read.table(loc_files[i], sep = ',', header = TRUE, stringsAsFactors = F)
  
  if (i == 1){
    all_tracks <- track
  } else{
    all_tracks <- rbind(all_tracks, track)
  }
  
}
 
all_tracks <- all_tracks[,c('deploymentID', 'time', 'argosLC', 'longitude', 'latitude')]
names(all_tracks) <- c('id','date','lc','lon','lat') # req'd names for fit_ssm
all_tracks$date <- lubridate::parse_date_time(all_tracks$date, orders = 'Ymd HMS', tz='UTC')
ssm_fit <- foieGras::fit_ssm(all_tracks, model = 'crw', time.step = 24, vmax = 4)
## this takes ~ 1 min on a Macbook Air with 1.6 GHz Intel Core i5 and 16 GB 2133 MHz LPDDR3

## throw an error if anything did not converge
if (any(!ssm_fit$converged)) stop('Not all models converged. Please check fit_ssm and try again.')

## grab predicted locations output from fit_ssm and group by id
plocs <- foieGras::grab(ssm_fit, what = "p", as_sf = FALSE) %>% 
  as_tibble() %>%
  mutate(id = as.character(id)) %>% group_by(id)

## take a look, for example, at # of standardized locations per individual
plocs %>% summarise(n = n())

## output level2 results
write.table(plocs, file = './data_level3/blue_sharks_level3.csv', sep = ',', col.names = TRUE, row.names = FALSE)

head(plocs)

```


Here's an example of the fits for one of the individuals.
```{r level3_example, eval = TRUE, message = FALSE, echo = FALSE}

ex_fit <- foieGras::fit_ssm(all_tracks[which(all_tracks$id == all_tracks$id[1]),], model = 'crw', time.step = 24, vmax = 4)

plot(ex_fit, what = 'predicted', type = 2)
#plot(ex_fit, what = 'f', type = 1)

```


## Level 3 -> 4

This step converts the time-standardized, regularized tracking data to grids that are spatially and temporally summarized. This example uses the count of observations per monthly, 1x1 degree grid cell.

```{r level3to4, eval = TRUE}

## grid to 1 month and 1deg x 1 deg as an example

## read level3 data
locs <- read.table('./data_level3/blue_sharks_level3.csv', sep = ',', header = TRUE)
locs$date <- lubridate::parse_date_time(locs$date, orders = c('Ymd HMS'), tz = 'UTC')
locs$month <- lubridate::month(locs$date)

## built base raster as template for monthly grids
ex <- raster::extent(min(locs$lon) - 2, max(locs$lon) + 2,
                     min(locs$lat) - 2, max(locs$lat) + 2)
r0 <- raster::raster(ex, res = 1)

## iterate by month to built a raster brick of monthly gridded counts per cell
for (i in 1:12){
  ## this summarizes the observation count per cell for month_i
  r <- raster::rasterize(cbind(locs$lon[which(locs$month == i)], locs$lat[which(locs$month == i)]), r0, fun = 'count')
  
  if (i == 1){
    ## create the output brick if month == 1
    month_grids <- raster::brick(r)
  } else{
    ## otherwise add to the existing output brick
    month_grids <- raster::addLayer(month_grids, r)
  }
}

writeRaster(month_grids, './data_level4/blue_shark_month_grids.grd', overwrite = TRUE)

```


```{r plot_grids, eval = TRUE, fig.width=12, fig.asp=1, echo = FALSE}
names(month_grids) <- format(seq(as.Date('2000-01-01'), as.Date('2000-12-01'), by='month'), '%b')
par(mfrow=c(4,3))
for (i in 1:12){
  plot(month_grids[[i]], main=names(month_grids[[i]]))
  world(add = T, fill = TRUE)

}

par(mfrow=c(1,1))
plot(sum(month_grids, na.rm = T), main='Sum of counts')
world(add = T, fill = TRUE)


```



